A Certified Interpreter for the List Machine Benchmark

### 2020-06-09 papers/SBLP2020/paper.tex
Small changes

\begin{itemize}
  \item We show how all the details of the list machine type system
        can be encoded as dependently-typed syntax which avoids, by construction,
        the presence of stuck states in its definitional interpreter.
  \item We provide a provably correct implementations for testing the subtyping
        relation and to calculate the least common super type of two input
        types for the machine registers.
\end{itemize}

@@

Each program variable is assigned a list type which is used to guarantee the safety of execute
fetch-field that demands non-empty list arguments and branch instructions refine the argument type
according to which branch is taken. In order to express such refinements, types are subject to a
subtyping relation. We let the meta-variable $\tau$ denote an arbitrary type. 
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
We let notation $\tau \subset \tau'$
denote the subtyping judgement which is defined as follows.
\[
\begin{array}{ccc}
  \inference{}
            {\tau\subset \tau}
            [refl]
  &
  \inference{}
            {nil \subset \tau}
            [nil]
  &
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [list]\\\\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [listcons]} \\\\
            \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [mixed]} \\\\
\end{array}
\]
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.

\begin{equation}
\inference{}
          {(r, (\iota_1;\iota_2);\iota_3) \xmapsto{p} (r, \iota_1;(\iota_2;\iota_3))}[step-seq]
\end{equation}

\begin{equation}
\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_0]=r'}
          {(r, (\textbf{fetch-field}~v~0~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-0]
\end{equation}

\begin{equation}
\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_1]=r'}
          {(r, (\textbf{fetch-field}~v~1~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-1]
\end{equation}

\begin{equation}
\inference{r(v_0) = a_0~~~r(v_1)=a_1~~~r[v':=cons(a_0,a_1)]=r'}
          {(r, (\textbf{cons}~v_0~v_1~v';\iota)) \xmapsto{p} (r', \iota)}[step-cons]
\end{equation}

\begin{equation}
\inference{r(v) = cons(a_0,a_1)}
          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota)}[step-branch-not-taken]
\end{equation}

\begin{equation}
\inference{r(v) = nil~~~p(l)=\iota'}
          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota')}[step-branch-taken]
\end{equation}

\begin{equation}
\inference{p(l)=\iota'}
          {(r, \textbf{jump}~l) \xmapsto{p} (r, \iota')}[step-jump]
\end{equation}

\begin{equation}
\inference{(r, \iota) \xmapsto{p} (r', \iota')~~~(p, r', \iota')}
          {(p, r, \iota) \Downarrow}[run-step]
\end{equation}

\begin{equation}
\inference{}
          {(p, r, \textbf{halt}) \Downarrow}[run-halt]
\end{equation}

\begin{equation}
\inference{\{\}[\textbf{v}_0:=\text{nil}]=r~~~p(\textbf{L}_0)=\iota~~~(p,r,\iota)\Downarrow}
          {p \Downarrow}[run-prog]
\end{equation}

%\subsection{Typing}

%\subsubsection{Subtyping}

\begin{equation}
\inference{}
          {\tau \subset \tau}[subtype-refl]
\end{equation}

\begin{equation}
\inference{}
          {\text{nil} \subset \text{list}~\tau}[subtype-nil]
\end{equation}

\begin{equation}
\inference{\tau \subset \tau'}
          {\text{list}~\tau \subset \text{list}~\tau'}[subtype-list]
\end{equation}

\begin{equation}
\inference{\tau \subset \tau'}
          {\text{listcons}~\tau \subset \text{list}~\tau'}[subtype-listcons]
\end{equation}

\begin{equation}
\inference{\tau \subset \tau'}
          {\text{listcons}~\tau \subset \text{listcons}~\tau'}[subtype-listmixed]
\end{equation}

>>>>

\begin{itemize}
  \item We show how all the details of the list machine type system
        can be encoded as dependently-typed syntax which avoids, by construction,
        the presence of stuck states in its definitional interpreter. In essence,
        the defined interpreter is a type soundness theorem for the
        list machine~\cite{Amin17}.
  \item We provide a provably correct implementations for testing the subtyping
        relation and to calculate the least common super type of two input
        types.
\end{itemize}

@@

Each program variable is assigned a list type which is used to guarantee the safety of execute
fetch-field that demands non-empty list arguments and branch instructions refine the argument type
according to which branch is taken. In order to express such refinements, types are subject to a
subtyping relation. We let the meta-variable $\tau$ denote an arbitrary type. 
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
We let notation $\tau \subset \tau'$
denote the subtyping judgement which is defined as follows.
\[
\begin{array}{ccc}
  \inference{}
            {\tau\subset \tau}
            [refl]
  &
  \inference{}
            {nil \subset \tau}
            [nil]
  &
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [list]\\\\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [listcons]} \\\\
            \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [mixed]} \\\\
\end{array}
\]
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
The other rules shows the that the $list$ and $listcons$ type constructors are
compatible with subtyping relation.
